---
title: "p8105_hw2_ys3637"
output: github_document
date: "2022-10-03"
---

```{r}
# library all packages that we need at the beginning
library(tidyverse)
library(dplyr)
library(readxl)
```

## Problem 1

This problem focuses on NYC Transit data; in particular, this CSV file contains information related to each entrance and exit for each subway station in NYC. If you’re not familiar with the NYC subway system, keeping a map in mind while looking at these data might help.

### a)

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).

```{r}
# read in data from CSV file
NYC_Transit <- read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
# show the first several lines of the original data
head(NYC_Transit)
# clean the data and convert "entry" from char to logical.
NYC_Transit <- NYC_Transit %>% 
  janitor::clean_names() %>% 
  select(line, station_name, station_latitude, station_longitude,
         route1:route11, entry, vending, entrance_type, ada) %>% 
  mutate(entry = recode(entry, 'YES' = TRUE, 'NO' = FALSE))
# show the first several lines of the cleaned data
head(NYC_Transit)
skimr::skim(NYC_Transit)
```

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

Answer:

This dataset contains 19 variables, including 11 character variables, e.g. line, 6 numeric variables, e.g. station latitude, station longitude, and 2 logical variables, e.g. ada, entry.

After I read in the data, I used clean_names() to make the all the variables name lowercase, and only select line, station_name, station_latitude, station_longitude, route, entry, vending, entrance_type, ada variables. Then I mutated the entry variable from character to logical type. 

There are 1868 rows and 19 columns. The data cannot be considered tidy, since the route variable spread over the table, and does not have its own column.

### b)

Answer the following questions using these data:

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.

```{r}
NYC_Transit %>% 
  select(station_name, line) %>% 
  distinct() %>% 
  count()
```

Answer: 465 distinct stations.

How many stations are ADA compliant?

```{r}
NYC_Transit %>% 
  select(station_name, line, ada) %>%
  filter(ada == TRUE) %>% 
  distinct() %>% 
  count()
```

Answer: 84 distinct stations are ADA compliant.

What proportion of station entrances / exits without vending allow entrance?

```{r}
# number of station without vending
Non_vending <-
  NYC_Transit %>% 
  select(station_name, line, vending) %>%
  filter(vending == "NO") %>% 
  distinct() %>% 
  count()
# proportion
Non_vending / 465
```

Answer: 21.29% proportion of station entrances / exits without vending allow entrance.

Reformat data so that route number and route name are distinct variables.
How many distinct stations serve the A train? 

```{r}
# converted route8-route11 to character variable
# made the route1 to route11 to be a new route_number variable while the value # under each route made a new value variable route_name
# deleted the prefix of route1 to route11 in order to only contain numbers
# removed all the rows which contains NA in route_name
NYC_Transit_tidy <- NYC_Transit %>% 
  mutate(route8 = as.character(route8), route9 = as.character(route9),
         route10 = as.character(route10), route11 = as.character(route11)) %>% 
  pivot_longer(route1:route11, names_to = "route_number",
               names_prefix = "route", values_to = "route_name") %>% 
  drop_na(route_name)

NYC_Transit_tidy %>% 
  select(station_name, line, route_name) %>%
  filter(route_name == "A") %>% 
  distinct() %>% 
  count()
```

Answer: 60 distinct stations serve the A train.

Of the stations that serve the A train, how many are ADA compliant?

```{r}
NYC_Transit_tidy %>% 
  select(station_name, line, route_name, ada) %>%
  filter(route_name == "A" & ada == TRUE) %>% 
  distinct() %>% 
  count()
```

Answer: Of the stations that serve the A train, 17 distinct stations are ADA compliant.

## Problem 2

This problem uses the Mr. Trash Wheel dataset, available as an Excel file on the course website.

Read and clean the Mr. Trash Wheel sheet:

specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel

use reasonable variable names

omit rows that do not include dumpster-specific data

round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

```{r}
# read in data from excel file
Mr_Trash <- read_excel("data/Trash Wheel Collection Data.xlsx",
                       sheet = 'Mr. Trash Wheel', range = 'A2:N550')
# show the first several lines of the original data
head(Mr_Trash)
# clean the data
Mr_Trash <- Mr_Trash %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(sports_balls = as.integer(round(sports_balls)))
# show the first several lines of the cleaned data
head(Mr_Trash)
```

Use a similar process to import, clean, and organize the data for Professor Trash Wheel, and combine this with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.

```{r}
# read in data from excel file
Professor_Trash <- read_excel("data/Trash Wheel Collection Data.xlsx",
                       sheet = 'Professor Trash Wheel', range = 'A2:M97')
# clean the data
Professor_Trash <- Professor_Trash %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(type_of_wheel = "Professor") %>% 
  select(type_of_wheel, everything())
# show the first several lines of the cleaned data
head(Professor_Trash)
# add the type of wheel to Mr_Trash
Mr_Trash <- Mr_Trash %>% 
  mutate(type_of_wheel = "Mr.", year = as.numeric(year)) %>% 
  select(type_of_wheel, everything())
# combine Mr_Trash and Professor_Trash
full_join(Mr_Trash, Professor_Trash)
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of sports balls collected by Mr. Trash Wheel in 2020?







